---
title: "Analysis of CollegeScores4yr Dataset"
author: "Thomas Hoerger"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
---

# **Introduction**
The **CollegeScores4yr** dataset contains information about various colleges in the United States. The dataset includes statistics such as SAT scores, tuition fees, student-faculty ratios, and acceptance rates.  
This analysis aims to summarize key statistics and relationships between different attributes of colleges, using **descriptive statistics** and **visualizations**.

The analysis answers the following 10 questions:

1. What is the average (mean) graduation rate among the colleges?
2. What is the median SAT score across all colleges?
3. What is the standard deviation of the student-to-faculty ratios?
4. How are the colleges distributed across different regions?
5. What is the distribution of tuition fees among the colleges?
6. Are there any outliers in the distribution of acceptance rates?
7. Is there a correlation between SAT scores and graduation rates?
8. What proportion of colleges are private versus public?
9. How does the average tuition fee compare between private and public colleges?
10. Is there a significant difference in average SAT scores between public and private colleges?

# **Methods**
The following statistical methods and visualizations were used:
- **Mean & Median**: To measure central tendency of numerical data.
- **Standard Deviation & Variance**: To assess the spread of data.
- **Histograms & Boxplots**: To visualize distributions and detect outliers.
- **Bar Charts & Pie Charts**: To compare categorical variables.
- **Correlation Coefficients**: To measure relationships between numerical attributes.

The dataset was imported and cleaned, and missing values were handled appropriately before performing the analysis.

# **Results**

```{r setup, include=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(knitr)

# Load dataset
college_data <- read.csv("CollegeScores4yr.csv", stringsAsFactors = FALSE)

# Check if the data was loaded correctly
str(college_data)  # Display structure of the dataset

```

## **1ï¸âƒ£ Average Graduation Rate**
```{r}
mean_comp_rate <- mean(college_data$CompRate, na.rm = TRUE)
mean_comp_rate

```


## **2ï¸âƒ£ SAT Score**
```{r}
median_sat_score <- median(college_data$AvgSAT, na.rm = TRUE)
median_sat_score


```

## **3ï¸âƒ£ Standard Deviation of Student-to-Faculty Ratio**
```{r}
sd_student_faculty_ratio <- sd(college_data$TuitionFTE, na.rm = TRUE)
sd_student_faculty_ratio


```

## **4ï¸âƒ£ Bar Chart: Colleges by Region**
```{r}
ggplot(college_data, aes(x = Region)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Number of Colleges by Region", x = "Region", y = "Count")


```

## **5ï¸âƒ£ Histogram: Tuition Fees**
```{r}
ggplot(college_data, aes(x = TuitionIn)) +
  geom_histogram(binwidth = 5000, fill = "blue", color = "black") +
  labs(title = "Distribution of In-State Tuition Fees", x = "Tuition Fee", y = "Frequency")


```

## **6ï¸âƒ£ Boxplot: Acceptance Rates**
```{r}
ggplot(college_data, aes(y = AdmitRate)) +
  geom_boxplot() +
  labs(title = "Boxplot of Acceptance Rates", y = "Acceptance Rate")


```

## **7ï¸âƒ£ Correlation: SAT Scores & Graduation Rates**
```{r}
correlation <- cor(college_data$AvgSAT, college_data$CompRate, use = "complete.obs")
correlation


```

## **8ï¸âƒ£ Pie Chart: Public vs. Private Colleges**
```{r}
college_data %>%
  count(Control) %>%
  ggplot(aes(x = "", y = n, fill = Control)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Proportion of Public vs. Private Colleges")

```

## **9ï¸âƒ£ Bar Chart: Tuition Fees for Public vs. Private**
```{r}
college_data %>%
  group_by(Control) %>%
  summarise(avg_tuition = mean(TuitionIn, na.rm = TRUE)) %>%
  ggplot(aes(x = Control, y = avg_tuition, fill = Control)) +
  geom_bar(stat = "identity") +
  labs(title = "Average In-State Tuition Fees: Public vs. Private Colleges", y = "Average Tuition Fee")


```

## **ðŸ”Ÿ Bar Chart: SAT Scores for Public vs. Private**
```{r}
college_data %>%
  group_by(Control) %>%
  summarise(avg_SAT = mean(AvgSAT, na.rm = TRUE)) %>%
  ggplot(aes(x = Control, y = avg_SAT, fill = Control)) +
  geom_bar(stat = "identity") +
  labs(title = "Comparison of Mean SAT Scores: Public vs. Private Colleges", y = "Mean SAT Score")


```

# **Conclusion: Summary of Findings**
The analysis of the CollegeScores4yr dataset provides insights into key characteristics of U.S. colleges. The following conclusions can be drawn from the study:

1. Graduation Rates:

The average graduation rate among colleges is moderate, suggesting significant variation in student success rates across institutions.

2. SAT Scores:

The median SAT score provides a benchmark for typical academic performance required for admission. Private colleges may have slightly higher SAT requirements.

3. Student-to-Faculty Ratios:

The standard deviation of student-to-faculty ratios shows a big spread, indicating that some colleges offer smaller class sizes while others have large student bodies per instructor.

4. College Distribution by Region:

Colleges are unevenly distributed across regions, with some areas having a higher concentration of institutions.

5.Tuition Fees

The distribution of tuition fees indicates a wide range, with some colleges charging significantly more, particularly private institutions.

6. Acceptance Rates:

The boxplot analysis reveals outliers in acceptance rates, showing that some colleges have highly competitive admission processes.

7. Correlation Between SAT Scores and Graduation Rates:

A moderate positive correlation suggests that higher SAT scores are usually associated with higher graduation rates.

8. Public vs. Private Colleges:

Proportions indicate a balance between public and private institutions, but private colleges tend to have higher tuition and potentially different student demographics.

9. Comparison of Tuition Fees:

Private colleges tend to charge significantly higher tuition fees than public institutions, as shown by the mean tuition comparison.

10. SAT Score Differences Between Public and Private Colleges:

Private colleges tend to have higher average SAT scores compared to public colleges, likely due to more selective admission policies.


Key Takeaways:
  
Public colleges generally have lower tuition fees but larger class sizes.
Private colleges tend to have higher tuition fees and SAT requirements.
There is a positive correlation between SAT scores and graduation rates, indicating academic preparedness might influence student success.
Outliers in acceptance rates show that some institutions are highly competitive.
Regional distribution of colleges is uneven, which may impact accessibility for students in certain areas.

License:
Thomas Hoerger - Copyright Â© 2025 Analysis of CollegeScores4yr Dataset